Bayesian Deep Gaussian Processes for Correlated Functional Data: \\ A Case Study in Cosmological Power Spectra

### 2025-04-04 paper/main.tex
move url to footnote?
edits to manuscript; include CAMB/higdon references, initial CAMB image, other small tweaks

We are particularly motivated by the Mira-Titan simulation suite \citep{moran2023mira}
which defines ``cosmologies'' according to eight cosmological parameters (more on this
in Section TBD).  For a particular cosmology, the simulation of the power matter spectra 
is rather complex.  Although we presume there exists some true matter spectra as a function
of wavenumber, we are not able to directly observe it.  Rather, the simulation outputs 
a convoluted patchwork of potential spectra, which vary in fidelity and which are trusted
only over particular wavenumber ranges.  Our first objective is to synthesize expert knowledge
and all of this simulation information into an effective estimate of the true underlying
matter spectrum.
\textcolor{red}{(@Steve - it might be worth moving something like Figure 1
here to help in this narration.  It is important that we impress upon the reader why
it is difficult to estimate the truth.  We can still refer back to the figure in 
Section 2 and narrate it further.)}

@@

Our second objective is to leverage simulation data from a limited set of
cosmologies in order to predict power matter spectra for unobserved cosmologies.
Due to the expense and propriatary nature of the Mira-Titan simulation suite
\textcolor{red}{(do we know how long it takes to run?  Is it really propriatary?)},
it is infeasible to evaluate the simulation for every possible eight-dimensional
cosmological configuration which may be of interest.  Instead, we desire
a statistical ``emulator'' or ``surrogate model'' 
\citep{santner2003design,gramacy2020surrogates} which will provide quick and effective
predictions of power matter spectra for any cosmological parameterization.
\citet{moran2023mira} provide a state-of-the-art emulator for the Mira-Titan
simulation suite, which was trained on 117 simulations, and is termed 
the ``Cosmic Emu''.\footnote{\url{https://github.com/lanl/CosmicEmu}}
We leverage the same training data, in conjunction with our Bayesian hierarchical
DGP model and basis function representations, to train a Gaussian process surrogate 
on the principal component weights in order to predict spectra for unobserved
cosmologies.  \textcolor{red}{(need more cites, like to Daves work)}
Our method compares favorably to Cosmic Emu on held-out cosmologies.

\textcolor{red}{(Should we introduce the CAMB data here?  I think so, if we
want to refer to it in the next paragraph.)}

The remainder of the paper is organized as follows.  Section \ref{sec:data} 
introduces our motivating application, the Mira-Titan simulation suite.  
Section \ref{sec:hm_fit} describes our hierarchical Bayesian DGP model and 
validates its performance on synthetic exercises and the CAMB simulation.  
Leveraging this trained model, Section \ref{sec:pred} details 
our procedure for predicting at unobserved cosmologies, benchmarking against
state-of-the-art competitors on the CAMB and Mira-Titan simulations. 
Section \ref{sec:disc} concludes with a discussion of our contributions and avenues for future research.  We provide reproducible code and an R package for our Bayesian DGP model
at \url{https://github.com/stevewalsh124/dgp.hm}. 
\textcolor{red}{Please note: this repository is currently private until we 
hear back regarding permissions for sharing the Mira-Titan data alongside our code.}

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=6in]{plot_data.png}
    \caption{Left: The perturbation theory, low resolution runs, and high resolution run for the first training cosmology. The weighted average of these is shown as a dashed line. Dotted lines at the bottom indicate where each data type is deemed approximately unbiased. Right: The same image, but restricted to $k$ values where the low resolution runs are approximately unbiased.}
    \label{fig:plot_data}
\end{figure}

@@

where $d=||x_i-x_j||^2$. From here, we estimate $\tau_\ell, \theta_\ell$ with maximum likelihood, and conditional on $\Sigma_\varepsilon$, we specify priors for $\tau_S, \theta_S$ following \cite{sauer2023active}: \textcolor{cyan}{Annie, please ensure the most up-to-date prior specs are shown above and here. Here is what I found from \texttt{fit\$settings} when using \texttt{deepgp v1.1.3}: $\theta_S \sim G(\alpha=1.2, \beta=1)$ and $\tau_S \propto 1/\tau_S^2$.}

@@

For the two variance settings, we consider matrices $\Sigma_4$ and $\Sigma_5$ to have a Mat\'ern covariance function with a nugget of $10^{-8}$ and smoothness of 2.5. For setting ``4", we have $\tau^2=0.0225$ and $\theta=0.01$. For setting ``5", we we have $\tau^2=0.01$ and $\theta=0.05$, and additionally scale this covariance matrix by a vector of values $s$ so that $\Sigma_5 = s M s^T$ is nonstationary; here $M$ is a Mat\'ern covariance matrix with parameters $\tau^2$ and $\theta$.

@@

In addition to simulated data, we investigate model fits the for different cosmological datasets (based on linear theory and/or runs from the Code for Anisotropies in the Microwave Background, CAMB). \textcolor{magenta}{We have looked at two sets of model runs, ``LT" runs where there is an underlying ``true" value alongside 15 computer model runs, or the ``CAMB" runs, where we only have a single output. The performance from the CAMB runs was better (with a 6-dimensional design space), and also did not require obtaining a posterior mean from multiple model runs. When looking at the CAMB runs on the $\Delta$ space, we see it generally is withing $\pm 5\%$ error. Regarding these results, compared to the results for the 8-dimensional LT runs which had less impressive predictive performance, I'm open to suggestions on whether to include details on one or the other.}

@@

\begin{figure}
    \centering
    %\includegraphics[width=\textwidth]{mean_PCs_oneW.png}
    \caption{The estimated mean trend of the 111 posterior means (left), the principal components contained in $B$ (middle), and an illustration of how the weights for the first PC will change as the first cosmological parameter varies (right) \textcolor{red}{This last plot is wonky because it was made with only the 6 test cosmologies, not a large uniform or FF design, Steve can update this}.}
    \label{fig:mean_PCs_oneW}
\end{figure}

@@

% \begin{figure}[ht]
%     \centering
%     \includegraphics[width=6in]{rmse_box.png}
%     \caption{Boxplot of all RMSEs of each method across all six test models. From top to bottom: box plots for our training method, process convolution training method, Cosmic EMU predictions, and our DGP-PC predictions.}
%     \label{fig:plot_rmse_box}
% \end{figure}

>>>>

We are particularly motivated by the Mira-Titan simulation suite \citep{moran2023mira}
which defines ``cosmologies'' according to eight cosmological parameters (\textcolor{red}{more on this
in Section TBD}).  For a particular cosmology, the simulation of the power matter spectra 
is rather complex.  Although we presume there exists some true matter spectra as a function
of wavenumber, we are not able to directly observe it.  Rather, the simulation outputs 
a convoluted patchwork of potential spectra, which vary in fidelity and which are trusted
only over particular wavenumber ranges.  Our first objective is to synthesize expert knowledge
and all of this simulation information into an effective estimate of the true underlying
matter spectrum.

\begin{figure}[ht]
    \centering
    \includegraphics[width=6in]{plot_data.png}
    \caption{Left: The perturbation theory, low resolution runs, and high resolution run 
             for the first training cosmology. The weighted average of these is shown as a dashed line. 
             Dotted lines at the bottom indicate where each data type is deemed approximately unbiased. 
             Right: The same image, but restricted to wavenumber ($k$) values where the low resolution 
             runs are approximately unbiased.}
    \label{fig:plot_data}
\end{figure}

@@

Our second objective is to leverage simulation data from a limited set of
cosmologies in order to predict power matter spectra for unobserved cosmologies.
Due to the computationally expensive nature of the Mira-Titan simulation suite
\textcolor{red}{(do we know how long it takes to run?)},
it is infeasible to evaluate the simulation for every possible eight-dimensional
cosmological configuration which may be of interest.  Instead, we desire
a statistical ``emulator'' or ``surrogate model'' 
\citep{santner2003design,gramacy2020surrogates} which will provide quick and effective
predictions of power matter spectra for any cosmological parameterization.
\citet{moran2023mira} provide a state-of-the-art emulator for the Mira-Titan
simulation suite, which was trained on 117 simulations, and is termed 
the ``Cosmic Emu''.\footnote{\url{https://github.com/lanl/CosmicEmu}}
We leverage the same training data, in conjunction with our Bayesian hierarchical
DGP model and basis function representations, to train a Gaussian process surrogate 
on the principal component weights in order to predict spectra for unobserved
cosmologies \citep{higdon2008computer, higdon2010estcosmo}. 
Our method compares favorably to Cosmic Emu on held-out cosmologies.

For a comprehensive assessment of our Bayesian hierarchical DGP model, we 
further study its performance in a simulation study with synthetic data mimicing 
the Mira-Titan data. Additionally, we also train our model with data output from 
the Code for Anisotropies in the Microwave Background \citep[CAMB,][]{lewis2011CAMB};
this model output exhibits similar behavior to Mira-Titan as well, but has the benefit 
of including a ``true" power spectrum corresponding toeach cosmology.

The remainder of the paper is organized as follows.  Section \ref{sec:data} 
introduces our motivating application, the Mira-Titan simulation suite.  
Section \ref{sec:hm_fit} describes our hierarchical Bayesian DGP model and 
validates its performance on synthetic exercises and the CAMB simulations.  
Leveraging this trained model, Section \ref{sec:pred} details 
our procedure for predicting at unobserved cosmologies, benchmarking against
state-of-the-art competitors on the CAMB and Mira-Titan simulations. 
Section \ref{sec:disc} concludes with a discussion of our contributions and avenues for 
future research.  We provide reproducible code and an R package for our Bayesian DGP model
in a public git repository.\footnote{\url{https://github.com/stevewalsh124/dgp.hm}}
\textcolor{red}{Please note: this repository is currently private until we 
hear back regarding permissions for sharing the Mira-Titan data alongside our code.}

@@

@@

where $d=||x_i-x_j||^2$. From here, we estimate $\tau_\ell, \theta_\ell$ with maximum 
likelihood, and conditional on $\Sigma_\varepsilon$, we specify priors for $\tau_S, \theta_S$ 
following \cite{sauer2023active}: \textcolor{cyan}{Annie, please ensure the most up-to-date 
prior specs are shown above and here. Here is what I found from fit\$settings when using 
deepgp v1.1.3: $\theta_S \sim G(\alpha=1.2, \beta=1)$ and $\tau_S \propto 1/\tau_S^2$.}

@@

For the two variance settings, we consider matrices $\Sigma_4$ and $\Sigma_5$ to 
have a Mat\'ern covariance function with a nugget of $10^{-8}$ and smoothness of 2.5. 
For setting ``4", we have $\tau^2=0.0225$ and $\theta=0.01$. For setting ``5", we have 
$\tau^2=0.01$ and $\theta=0.05$, and additionally scale this covariance matrix by 
a vector of values $s$ so that $\Sigma_5 = s M s^T$ is nonstationary; here $M$ is 
a Mat\'ern covariance matrix with parameters $\tau^2$ and $\theta$.

@@

In addition to simulated data, we investigate model fits for different cosmological 
datasets from CAMB \citep{lewis2011CAMB}. \textcolor{magenta}{We have looked at two sets of model 
runs, ``LT" runs where there is an underlying ``true" value alongside 15 computer 
model runs, or the ``CAMB" runs, where we only have a single output. The performance 
from the CAMB runs was better (with a 6-dimensional design space), and also did not 
require obtaining a posterior mean from multiple model runs. When looking at the 
CAMB runs on the $\Delta$ space, we see it generally is withing $\pm 5\%$ error. 
Regarding these results, compared to the results for the 8-dimensional LT runs 
which had less impressive predictive performance, I'm open to suggestions on whether 
to include details on one or the other.}

\begin{figure}
    \centering
    \includegraphics[width=6in]{CAMB_fit_model1.jpg}
    \caption{Example of model fit for the first model from the CAMB data, with LOESS
             mean term subtracted. \textcolor{red}{Steve: Update this plot to be
             more clean, use different line types: dotted, dash, solid, etc.}}   
    \label{fig:fit_camb}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{mean_PCs_oneW.png}
    \caption{The estimated mean trend of the 111 posterior means (left), the principal 
    components contained in $B$ (middle), and an illustration of how the weights 
    for the first PC will change as the first cosmological parameter varies (right) 
    \textcolor{red}{This last plot is wonky because it was made with only the 6 test 
    cosmologies, not a large uniform or FF design, Steve can update this}.}
    \label{fig:mean_PCs_oneW}
\end{figure}

@@

% \begin{figure}[ht]
%     \centering
%     \includegraphics[width=6in]{rmse_box.png}
%     \caption{Boxplot of all RMSEs of each method across all six test models. 
%.             From top to bottom: box plots for our training method, process 
%              convolution training method, Cosmic EMU predictions, and our DGP-PC predictions.}
%     \label{fig:plot_rmse_box}
% \end{figure}
